---
title: "Wine Evaluation"
author: "John Cruz, Noori Selina, Shaya Engelman, Daniel Craig, Gavriel Steinmetz-Silber"
date: "2024-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message = FALSE)


# include = FALSE prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
# echo = FALSE prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
# message = FALSE prevents messages that are generated by code from appearing in the finished file.
# warning = FALSE prevents warnings that are generated by code from appearing in the finished.
# fig.cap = "..." adds a caption to graphical results.
```

## Required Libraries

```{r library,include = TRUE, class.source = "fold-show"}
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(latex2exp)
library(psych)
library(scales)
library(stringr)
library(ggcorrplot)
library(ggmice)
```

To-Do List:
1. Check for typo's - No Typo's, all data is int
2. Check for missing
3. Show distributions
4. Determine Categorical/Continuous


```{r Load Data}

url <- "https://raw.githubusercontent.com/d-ev-craig/DATA621_Group/main/HW5/data/wine-training-data.csv"
eval_url <- "https://raw.githubusercontent.com/d-ev-craig/DATA621_Group/main/HW5/data/wine-evaluation-data.csv"

train <- read_csv(url)
eval <- read_csv(eval_url)

train <- train %>% select(-INDEX)
```

# Introduction

A data set containing information on approximately 12,000 commercially available wines and their variables mostly related to  chemical properties is analyzed for impact on sales and used to predict on sales to give accurate forecasts for manufacturing. The response variable is the number of sample cases of wine that were purchased by wine distribution companies after sampling a wine. These cases would be used to provide tasting samples to restaurants and wine stores around the United States. The more sample cases purchased, the more likely is a wine to be sold at a high end restaurant.

## Data Summary

A table below expands on the variables included in analysis with comments from domain experts on expected effects. 

```{r}
summary <- data.frame('Variable_Name' = c('INDEX','TARGET','AcidIndex', 'Alcohol','Chlorides','CitricAcid','Density','FixedAcidity','FreeSulfurDioxide','LabelAppeal','ResidualSugar','STARS','Sulphates','TotalSulfurDioxide','VolatileAcidity','pH'),
                      'Description' = c('Identification Variable','Number of Cases Purchased', 'Proprietary Method of Testing Total Acidity','Alcohol Content','Chloride Content of Wine','Citric Acid Content', 'Density of Wine', 'Fixed Acidity of Wine', 'Sulfur Dioxide Content of Wine', 'Marketing Score indicating the appeal of label design for consumers. High numbers 
suggest customers like the label design.','Residual Sugar of Wine','Wine Rating by a Team of Experts, 4 Stars = Excellent','Sulfate Content of Wine', 'Sulfur Dioxide of Wine', 'Volatile Acid Content of Wine', 'pH of wine'),
                      'Theoretical_Effect' = c('','','','','','','','','','Many consumers purchase based on the visual appeal of the wine label design. Higher numbers suggest better sales.','','A high number of stars suggests high sales.','','','',''),
                      'Variable_Type' = c('N/A','Discrete','Continuous','Continuous','Continuous','Continuous','Continuous','Continuous','Continuous','Ordinal','Continuous','Ordinal','Continuous','Continuous','Continuous','Continuous'))


kbl(summary, caption = "Summary") %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(latex_options = "HOLD_position")

```

A quick look at the variables 5 number summary reveals that several variables have large ranges which when relating to their mean may suggest significantly different scales between variables, a high amount of skew, bi-modal distributions, or outliers. FixedAcidity, ResidualSugar, FreeSulfurDioxide, and TotalSulfurDioxide have fairly extreme ranges in comparison to their means.  Variables with Kurtosis greater than 4 will have observations distributed into heavy or long tails and may suggest numerous outliers, less than 2 suggest distributions centered around their mean with short or thin tails. Many of the variables are just below 2 suggesting many will have sharp peaks around the mean. Only AcidIndex shows as a non-ordinal or discrete distribution with extreme values of kurtosis, suggesting it will contain many outliers. 
```{r}
desc_train <- train %>% describe(omit = TRUE) %>% select(-vars, -n)

desc_train |> 
  rownames_to_column('vars') %>% # used to store row names (mutate deletes them)
  mutate(
  range = cell_spec(range, color = ifelse(range > 50, "white", "black"),
                  background = ifelse(range > 50, "red", "white"),
                  bold = ifelse(range > 50, T, F)),
  skew= cell_spec(skew, color = ifelse(skew > 1 | skew < -1, "white", "black"),
                  background = ifelse(skew > 1 | skew < -1, "red", "white"),
                  bold = ifelse(skew > 1 | skew < -1, T, F)),
  kurtosis = cell_spec(kurtosis, color = ifelse(kurtosis > 4 | kurtosis < 1, "white", "black"),
                  background = ifelse(kurtosis > 4 | kurtosis < 1, "red", "white"),
                  bold = ifelse(kurtosis > 4 | kurtosis < 1, T, F))
  ) %>%
  column_to_rownames('vars') %>% # used to put row names back in place
  
  kable(caption = "5 Number Summary",format = "html",escape = F, booktabs = T) %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(latex_options = "HOLD_position")



```

**Histograms**

As kurtosis foreshadowed, many of the distributions have sharp peaks at the mean with only the AcidIndex showing a bi-modal distribution. With the sharp centers around the peaks in the histograms, a high number of outliers may present themselves.
```{r histograms, echo=FALSE, warning=FALSE}
train %>% select(-c(LabelAppeal,STARS,TARGET))|>
  gather(key = "variable", value = "value") |>  
  ggplot(aes(x = value)) + 
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = '#4E79A7', color = 'black') + 
  stat_density(geom = "line", color = "red") +
  facet_wrap(~ variable, scales = 'free') +
  theme(strip.text = element_text(size = 5)) +
  theme_bw()
```

** Bar Plots **

There is a relatively normal distribution to LabelAppeal, but both STARS and TARGET tend to favor their lower values suggesting it's quite difficult to gain either a critic's praise or a significant amount of cases sold.
```{r barplots}
train %>% select(c(LabelAppeal,STARS,TARGET)) |>
  gather(key = "variable", value = "value") |>  
  ggplot(aes(x = value)) + 
  geom_bar(aes(y = after_stat(count)), bins = 20, fill = '#4E79A7', color = 'black') +
  facet_wrap(~ variable, scales = 'free') +
  theme(strip.text = element_text(size = 5)) +
  theme_bw() +
  labs(y = "") 
```

Boxplots reveal a significant number of residuals in all of the variables.
```{r boxplot}
train %>% select(-c(LabelAppeal,STARS,TARGET)) %>%
  gather(key = "Variable", value = "Value") |>
  ggplot(aes(x = "", y = Value)) +  
  geom_boxplot(fill = "#4E79A7") +
  facet_wrap(~ Variable, scales = "free") + 
  labs(x = NULL, y = "Value") +  
  theme(strip.text = element_text(size = 5))
```

**Correlation Matrix**

The correlation matrix reveals a moderate relationship between STARS and LabelAppeal with Target. Although both STARS and LabelAppeal seem to be somewhat correlated to each other suggesting potential colinearity. The AcidIndex, being a propietary method that aggregates across Acid metrics, does show some relationship with FixedAcidity but is relatively minor.

```{r corr-plot}
q <- cor(train, use = 'complete.obs')

ggcorrplot(q, type = "upper", outline.color = "white",
           ggtheme = theme_classic,
           colors = c("#F28E2B", "white", "#4E79A7"),
           lab = TRUE, show.legend = F, tl.cex = 5, lab_size = 3) 
```

**Missing Values**

While missing values may be indicative of the target, the STARS variable is missing 26% of its values. Determining the relationship it has to cases sold may be useful before removing it from the dataset. Since a value of zero is not given to any wine, and we wish to see the relationship of these no star wines to their cases sold, the missing values were replaced with zero. Chlorides, FreeSulfurDioxide, Alcohol, and TotalSulfurDioxide are missing around 5% or about 600 values. Sulphates is missing about 10% of its values and about 1200 values.
```{r percent missing check}
percentMiss <- function(x){sum(is.na(x))/length(x)*100} # Creates percentage of missing values

# Cut offs for variable dropping was 25% of values missing - none were dropped
# Cut offs for sample dropping was 50% of values missing - none were dropped

variable_pMiss <- apply(train,2,percentMiss) # 2 = runs on columns
sample_pMiss <- apply(train,1,percentMiss) # 1 = runs on rows

#sum(sample_pMiss > 50) 

pMiss <- data.frame(variables = names(variable_pMiss),pMiss = (variable_pMiss), row.names = NULL)
pMiss <- pMiss %>% arrange(desc(pMiss))


pMiss |>
  ggplot(aes(x = reorder(variables,pMiss), y = pMiss)) + 
  geom_bar(stat = 'identity', fill = '#4E79A7', color = 'black') +
  theme(strip.text = element_text(size = 5)) +
  theme_bw() +
  scale_x_discrete(guide = guide_axis(angle = 45))+
  labs(x = 'Variables',y = 'Percent Missing',title = 'Percent of Missing Values by Variable')

```

```{r missing-values, echo=FALSE}
missing_val <-
  train %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  select_if(function(.) last(.) != 0)

kbl(missing_val, caption = "Missing Values Count") |>
    kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(latex_options = "HOLD_position")
```

```{r missing pattern, echo=TRUE}
plot_pattern(train, square = TRUE, rotate = TRUE, npat = 6)

```

While non-rated wines (shown below as a value of 0 stars assigned) typically aren't purchased, there are some that sell about 3 cases. This might suggest that non-rated wines are not submitted for critic's appraisal and should be used as a feature in the modeling. This plot also reveals a heavy preference for 2 star wines.
```{r}
train$STARS <- as.factor(train$STARS)
train$TARGET <- as.factor(train$TARGET)
train$LabelAppeal <- as.factor(train$LabelAppeal)

library(ggmosaic)
library(viridis)

ggplot(train, aes(STARS, TARGET)) + 
  scale_fill_viridis() +
  geom_bin2d()+
  labs(title = "Cases Sold vs STARS",
       subtitle = "STARS: 4 = Excellent; 1 = Poor")
```


The majority of 0 value appeals center on 4 cases sold and does tend to show a linear relationship between the two.
```{r}
ggplot(train, aes(LabelAppeal, TARGET)) + 
  scale_fill_viridis() +
  geom_bin2d()+
  labs(title = "Cases Sold vs LabelAppeal",
       subtitle = "LabelAppeal: 2 = High Appeal; -2 = Low Appeal")
```


